---
title: "R and R Studio appendi exercises"
author: "Nicholas Horton (nhorton@amherst.edu)"
date: "August 17, 2017"
output:
  html_document:
    fig_height: 5
    fig_width: 7
  pdf_document:
    fig_height: 5
    fig_width: 7
  word_document:
    fig_height: 3
    fig_width: 5
---

```{r, setup, include=FALSE}
library(mdsr)   # Load additional packages here 
library(tidyr)
library(tm)
library(wordcloud)


# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
```

## Introduction
These exercises are taken from the R and RStudio appendix from **Modern Data Science with R**: http://mdsr-book.github.io.  Other materials relevant for instructors (sample activities, overview video) for this chapter can be found there.


## Typing commands
A user has typed the following commands into the console.

```{r}
obj1 <- 2:10
obj2 <- c(2, 5)
obj3 <- c(TRUE, FALSE)
obj4 <- 42
```

What values are returned by the following commands?

```{r eval=FALSE}
obj1 * 10
obj1[2:4]
obj1[-3]
obj1 + obj2
obj1 * obj3
obj1 + obj4
obj2 + obj3
sum(obj2)
sum(obj3)
```

SOLUTION:

  

## Typing more commands
A user has typed the following commands into the console.

```{r}
a <- c(10, 15)
b <- c(TRUE, FALSE)
c <- c("happy", "sad")
```

What do each of the following commands return?  Describe the class of the object as well as its value.

```{r eval=FALSE}
data.frame(a, b, c)
cbind(a, b)
rbind(a, b)
cbind(a, b, c)
list(a, b, c)[[2]]
```

SOLUTION:


## Typing even more commands
A user has typed the following commands into the console.

```{r}
mylist <- list(x1="sally", x2=42, x3=FALSE, x4=1:5)
```

What values do each of the following commands return?
```{r eval=FALSE}
is.list(mylist)
names(mylist)
length(mylist)
mylist[[2]]
mylist[["x1"]]
mylist$x2
length(mylist[["x4"]])
class(mylist)
typeof(mylist)
class(mylist[[4]])
typeof(mylist[[3]])
```

SOLUTION:

## Piping
The following code undertakes some data analysis using the HELP (Health Evaluation and Linkage to Primary Care) trial.

```{r eval=FALSE}
library(mosaic)
ds <-
   read.csv("http://nhorton.people.amherst.edu/r2/datasets/helpmiss.csv")
summarise(group_by(select(filter(mutate(ds,
  sex = ifelse(female==1, "F", "M")), !is.na(pcs)), age, pcs, sex),
  sex), meanage=mean(age), meanpcs=mean(pcs),n=n())
```

Describe in words what computations are being done.
Using the 'pipe' notation, translate this code into a more readable version.

SOLUTION:

```{r}
# solution goes here
```

## Concepts
The following concepts should have some meaning to you: package, function, command, argument, assignment,
object, object name, data frame, named argument, quoted character string.  Construct an example of R commands
that make use of at least four of these.  Label which part of your example R command corresponds to each.

SOLUTION:


## Names
Which of these kinds of names should be wrapped with quotation marks when used in R?
1. function name
2. file name
3. the name of an argument in a named argument
4. object name

SOLUTION:


## What's wrong?
What's wrong with this statement?
```{r eval=FALSE}
help(NHANES, package <- "NHANES")
```


## CPS
Consult the documentation for `CPS85` in the `mosaicData` package to determine the meaning of CPS.

SOLUTION:

```{r}
library(mosaicData)
# solution does here
```


## Errors
For each of the following assignment statements, describe the error (or note why it does not generate an error).
```{r eval=FALSE}
result1 <- sqrt 10
result2 <-- "Hello to you!"
3result <- "Hello to you"
result4 <- "Hello to you
result5 <- date()
```

SOLUTION:

