---
title: "Grammar of graphics exercises"
author: "Nicholas Horton (nhorton@amherst.edu)"
date: "August 31, 2017"
output:
  html_document:
    fig_height: 5
    fig_width: 7
  pdf_document:
    fig_height: 5
    fig_width: 7
  word_document:
    fig_height: 3
    fig_width: 5
---

```{r, setup, include=FALSE}
library(mdsr)   # Load additional packages here 


# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
```

## Introduction
These exercises are taken from the grammar of graphics chapter from **Modern Data Science with R**: http://mdsr-book.github.io.  Other materials relevant for instructors (sample activities, overview video) for this chapter can be found there.


## Galton
Using the famous `Galton` data set from the `mosaicData` package:

```{r}
library(mdsr)
glimpse(Galton)
```

1. Create a scatterplot of each person's \var{height} against their father's height
2. Separate your plot into facets by \var{sex}
3. Add regression lines to all of your facets

Hint: recall that you can find out more about the data set by running the command `?Galton`.


SOLUTION:

```{r}
library(mdsr)
glimpse(Galton)
# solution goes here
```


## Railtrails
Using the `RailTrail` data set from the `mosaicData` package:

1.  Create a scatterplot of the number of crossings per day \var{volume} against the high temperature that day
2. Separate your plot into facets by \var{weekday}
3. Add regression lines to the two facets

SOLUTION:

```{r}
library(mdsr)
glimpse(RailTrail)
# solution goes here
```
