---
title: "Grammar of graphics exercises"
author: "Nicholas Horton (nhorton@amherst.edu)"
date: "August 31, 2017"
output:
  html_document:
    fig_height: 5
    fig_width: 7
  pdf_document:
    fig_height: 5
    fig_width: 7
  word_document:
    fig_height: 3
    fig_width: 5
---

```{r, setup, include=FALSE}
library(mdsr)   # Load additional packages here 


# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
```

## Introduction
These exercises are taken from the grammar of graphics chapter from **Modern Data Science with R**: http://mdsr-book.github.io.  Other materials relevant for instructors (sample activities, overview video) for this chapter can be found there.


## Galton
Using the famous `Galton` data set from the `mosaicData` package:

1. Create a scatterplot of each person's \var{height} against their father's height
2. Separate your plot into facets by \var{sex}
3. Add regression lines to all of your facets

Hint: recall that you can find out more about the data set by running the command `?Galton`.


SOLUTION:

```{r}
library(mdsr)
glimpse(Galton)
# solution goes here
```


## Railtrails
Using the `RailTrail` data set from the `mosaicData` package:

1.  Create a scatterplot of the number of crossings per day \var{volume} against the high temperature that day
2. Separate your plot into facets by \var{weekday}
3. Add regression lines to the two facets

SOLUTION:

```{r}
library(mdsr)
glimpse(RailTrail)
# solution goes here
```


## Hamilton and Angelica
Angelica Schuyler Church (https://en.wikipedia.org/wiki/Angelica_Schuyler_Church, 1756--1814) was the daughter of New York Governer Philip Schuyler and sister of
Elizabeth Schuyler Hamilton.  Angelica, New York was named after her.  Generate a plot of the reported proportion of babies born
with the name Angelica over time and interpret the figure.

SOLUTION:

```{r}
library(mdsr)
library(babynames)
glimpse(babynames)
# solution goes here
```

## Marriage
The following questions use the `Marriage` data set from the `mosaicData` package.

1.  Create an informative and meaningful data graphic.
2. Identify each of the visual cues that you are using, and describe how they are related to each variable.
3. Create a data graphic with at least \emph{five} variables (either quantitative or categorical). For the purposes of this exercise, do not worry about making your visualization meaningful---just try to encode five variables into one plot.


```{r}
library(mdsr)
glimpse(Marriage)
# solution goes here
```

## MLB teams
The `MLB_teams` data set in the `mdsr` package contains information about Major League Baseball teams in the past four seasons. There are several quantitative and a few categorical variables present. See how many variables you can illustrate on a single plot in R. The current record is 7. (Note: This is *not* good graphical practice---it is merely an exercise to help you understand how to use visual cues and aesthetics!)

```{r}
library(mdsr)
glimpse(MLB_teams)
```


