---
title: "Tidy data and iteration exercises"
author: "Nicholas Horton (nhorton@amherst.edu)"
date: "August 31, 2017"
output:
  html_document:
    fig_height: 5
    fig_width: 7
  pdf_document:
    fig_height: 5
    fig_width: 7
  word_document:
    fig_height: 3
    fig_width: 5
---

```{r, setup, include=FALSE}
library(mdsr)   # Load additional packages here 


# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
```

## Introduction
These exercises are taken from the tidy data and iteration chapter from **Modern Data Science with R**: http://mdsr-book.github.io.  Other materials relevant for instructors (sample activities, overview video) for this chapter can be found there.


## Home runs
Consider the number of home runs hit (`HR`) and home runs allowed (`HRA`) for the Chicago Cubs (CHN) baseball team. Reshape the `Teams` data from the `Lahman` package into *long* format and plot a time series conditioned on whether the HRs that involved the Cubs were hit by them or allowed by them.

SOLUTION:

```{r}
library(mdsr)
library(Lahman)
# solution goes here
```


## Seasons
Write a function called `count_seasons` that, when given a `teamID`, will count the number of seasons the team played in the `Teams` data frame from the `Lahman` package.

SOLUTION:

```{r}
library(mdsr)   
library(Lahman)
# solution goes here
```

## We'll always have Brooklyn
The team IDs corresponding to Brooklyn baseball teams from the `Teams` data frame from the `Lahman` package are listed below. Use `sapply()` to find the number of seasons in which each of those teams played.

SOLUTION:

```{r}
library(mdsr)   
library(Lahman)
bk_teams <- c("BR1", "BR2", "BR3", "BR4", "BRO", "BRP", "BRF")
# solution goes here
```

## Marriage
In the `Marriage` data set included in `mosaicData`, the `appdate`, `ceremonydate`, and `dob` variables are encoded as factors, even though they are dates. Use the `lubridate` package to convert those three columns into a date format.

SOLUTION:

```{r}
library(mdsr)   
Marriage %>%
  select(appdate, ceremonydate, dob) %>%
  glimpse()
# solution goes here
```

## Coercion
Consider the values returned by the `as.numeric()` and `readr::parse_number()` functions when applied to the following vectors.  Describe the results and their implication.

<<>>=
x1 <- c("1900.45", "$1900.45", "1,900.45", "nearly $2000")
x2 <- as.factor(x1)
@


## Brittle code
An analyst wants to calculate the pairwise differences between the Treatment and Control values for a small data set from a crossover trial (all subjects received both treatments) that
consists of the following observations.
```{r echo=FALSE}
library(xtable)
ds1 <- data_frame(id=c(1:3, 1:3), group=c(rep("T", 3), rep("C", 3)), vals=c(4,6,8,5,6,10))
```
```{r results="asis"}
tab <- xtable(ds1)
print(tab, type="html", floating=FALSE)
```

They use the following code to create the new `diff` variable.

```{r eval=FALSE}
Treat <- filter(ds1, group=="T")
Control <- filter(ds1, group=="C")
all <- mutate(Treat, diff = Treat$vals - Control$vals)
all
```

Verify that this code works for this example and generates the correct values of -1, 0, and -2.  Describe two problems that might arise if the data set is not sorted in a particular
order or if one of the observations is missing for one of the subjects.  Provide an alternative approach to generate this
variable that is more robust (hint: use `tidyr::spread()`).

SOLUTION:

```{r message=FALSE}
# solution goes here
```

## Tall to wide
Generate the code to convert the following data frame to wide format.
```{r echo=FALSE}
library(mosaic)
library(readr)
library(tidyr)
df <- data_frame(
  id=c(1,1,2,2,3,3,4,4,5,5),
  sex=c("M","M","F","F","F","F","M","M","M","M"),
  grp=rep(c("A","B"), 5),
  left=c(0.1,0.2, 0.3,0.4, 0.15,0.25, 0.6,0.65, 0.71,0.79),
  right=c(0.1,0.25, 0.3,0.35, 0.18,0.25, 0.6,0.6, 0.71,0.79)+0.1
)
sumdf <- df %>%
  group_by(grp, sex) %>%
  summarize(meanL=mean(left), sdL=sd(left),
            meanR=mean(right), sdR=sd(right)) %>%
  ungroup()
```

```{r echo=FALSE, results='asis'}
tab <- xtable(sumdf, floating=FALSE)
print(tab, type="html")
```

The result should look like the following display.
```{r echo=FALSE, results="asis"}
transformed <- sumdf %>%
  gather(key = "temp", value = "vals", meanL, meanR, sdL, sdR) %>%
  unite(col = "temp1", sex, temp, sep = ".") %>%
  spread(temp1, vals)
tab <- xtable(data.frame(transformed))
print(tab, type="html")
```

Hint: use `gather()` in conjunction with `spread()`.

## Multiple models
Use the `dplyr::do()` function and the `HELPrct` data frame from the `mosaicData` package to fit a
regression model predicting `cesd` as a function of `age` separately for each of the levels
of the `substance` variable.  Generate a table of results (estimates and confidence intervals) for each level of the grouping variable.
New York to `SFO` using the `nycflights13` package.  How do the conclusions change from
the multiple regression model presented in the Statistical Foundations Chapter?

SOLUTION:

```{r}
library(mdsr)   
# solution goes here
```

## Baseball records
Use the `dplyr::do()` function and the `Lahman` data to replicate one of these baseball records plots (http://tinyurl.com/nytimes-records) from the *The New York Times*.

SOLUTION:

```{r}
library(mdsr)   
library(Lahman)
# solution goes here
```

## FEC
Use the `fec` package to download the Federal Election Commission data for 2012. Re-create Figures 2.1 and 2.2 using `ggplot2`.

SOLUTION:

```{r}
# solution goes here, after downloading the fec package.
```

## More FEC
Using the same FEC data as the previous exercise, re-create Figure 2.8.  

SOLUTION:

```{r}
# solution goes here
```

## Wikipedia
Using the approach described in Section 5.5.4, find another table in Wikipedia
that can be scraped and visualized.  Be sure to interpret your graphical display.

SOLUTION:

```{r}
library(mdsr)
# solution goes here
```

## Wrangling with the FEC
Replicate the wrangling to create the `house_elections` table in the `fec` package from the original Excel source file.

SOLUTION:

```{r}
library(mdsr)
# solution goes here
```

## Babynames
Replicate the functionality of the `make_babynames_dist()` function from the `mdsr` package to wrangle the original tables from the `babynames` package.

SOLUTION:

```{r}
library(mdsr)
# solution goes here
```
