---
title: "Text as Data exercises"
author: "Nicholas Horton (nhorton@amherst.edu)"
date: "July 19, 2017"
output:
  html_document:
    fig_height: 5
    fig_width: 7
  pdf_document:
    fig_height: 5
    fig_width: 7
  word_document:
    fig_height: 3
    fig_width: 5
---

```{r, setup, include=FALSE}
library(mdsr)   # Load additional packages here 
library(tidyr)
library(tm)
library(wordcloud)


# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
```

## Introduction
These exercises are taken from the text as data chapter from **Modern Data Science with R**): http://mdsr-book.github.io/.  Other materials relevant for instructors (sample activities, overview video) for this chapter can be found there.


## Speaking lines
Speaking lines in Shakespeare's plays are identified by a line that starts with two spaces, then a string of capital letters and spaces (the character's name) followed by a period. Use `grep()` to find all of the speaking lines in *Macbeth*. How many are there? 

SOLUTION:

```{r}
library(mdsr)   
library(tidyr)
library(tm)
library(wordcloud)
data(Macbeth_raw)
macbeth <- strsplit(Macbeth_raw, "\r\n")[[1]]
head(macbeth)
# solution goes here
```

  

## Hyphenated words
Find all the hyphenated words in one of Shakespeare's plays.

SOLUTION:

```{r}
# solution goes here
```



## Most popular names
Use the `babynames` data table from the `babynames` package to find the ten
most popular:

1) Boys' names ending in a vowel.

SOLUTION:

```{r}
# solution goes here
```

2) Names ending with `joe`, `jo` `Joe` or `Jo` (e.g., `Billyjoe`).

SOLUTION:

```{r}
# solution goes here
```

## Adjectives

Find all of the adjectives in one of Shakespeare's plays that end in `more` or `less` (note change from original question 15.4).

SOLUTION:

```{r}
# solution goes here
```

## Stage directions

Find all of the lines containing the stage direction \val{Exit} or \val{Exeunt} in one of Shakespeare's plays (note change from original exercise 15.5).

SOLUTION:

```{r}
# solution goes here
```

## Regular expressions
Use regular expressions to determine the number of speaking lines from the *Complete Works of William Shakespeare* (http://www.gutenberg.org/cache/epub/100/pg100.txt). Here, we care only about how many times a character speaks---not what they say or for how long they speak. 

SOLUTION:

```{r}
# solution goes here
```

  
## Top characters
Make a bar chart displaying the top 100 characters with the greatest number of lines. 
*Hint*: you may want to use either the `stringr::str_extract()` or `strsplit()` function here.

SOLUTION:

```{r}
# solution goes here
```


## Shakespare Machine
In this problem, you will do much of the work to recreate Mark Hansen's *Shakespeare Machine*. Start by watching a video clip (http://vimeo.com/54858820) of the exhibit.
Use *The Complete Works of William Shakespeare* (see earlier exercise) and regular expressions to find all of the hyphenated words in Shakespeare Machine. How many are there? 
Use `%in\%` to verify that your list contains the following hyphenated words pictured at 00:46 of the clip.


SOLUTION:

```{r}
sm_words <- c("true-fix'd", "pale-hearted", "lean-fac'd", "hard-hearted", 
  "best-regarded", "thick-ribbed", "both-sides", "sea-like.", 
  "shrill-shrieking", "lust-stain'd", "tragical-historical,")
# solution goes here
```


## Wikipedia table
Find an interesting Wikipedia page with a table, scrape the data from it, and generate a figure that tells an interesting story. 
Include an interpretation of the figure.

SOLUTION:

```{r}
# solution goes here
```


## Stackexchange 1
The site \url{stackexchange.com} displays questions and answers on technical topics.  
The following code downloads the most recent \R questions related to the \pkg{dplyr} package. 

```{r message=FALSE}
library(httr)
# Find the most recent R questions on stackoverflow
getresult <- GET("http://api.stackexchange.com",
  path = "questions",
  query = list(site = "stackoverflow.com", tagged = "dplyr"))
stop_for_status(getresult) # Ensure returned without error
questions <- content(getresult)  # Grab content
names(questions$items[[1]])    # What does the returned data look like?
length(questions$items)
substr(questions$items[[1]]$title, 1, 68)
substr(questions$items[[2]]$title, 1, 68)
substr(questions$items[[3]]$title, 1, 68)
```

How many questions were returned?
Without using jargon, describe in words what is being displayed and how it might be
used.

SOLUTION:

```{r}
# solution goes here
```



## Stackexchange 2
Repeat the process of downloading the content from \url{stackexchange.com} related to 
the \pkg{dplyr} package and summarize the results.


SOLUTION:

```{r}
# solution goes here
```




